---
title: HOME
description: Welcome to this sample project
images: ["/images/sample.jpg"]
---

Welcome to my personal website,

**This website is still under contruction.**

[Get to know me better](/resume "Get to know me better")

### Selecting Mutual Funds in Colombia

A lot of friends and colleagues usually ask me what they should do with their savings and where to invest them. This is not an easy question from a technical point of view and the answer depends on your risk aversion degree, your investment horizon and how much you have to invest, among other important variables. For that reason, I decided to tackle this question using information of all of the Colombian mutual funds available for the general public, using [sifi data](https://sificcolombia.lvaindices.com/estadisticas/).

*(Information provided here do not constitute any kind of financial/investment advice but is only an exercise for academic purposes)*

#### Overiview

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(readxl)

dataset <- 
  map_dfr(str_subset(dir("../static/sifi"), ".xlsx$"), 
          function(i) {
            df <- read_excel(str_c("../static/sifi/", i))[, 1:9]
            df[, 1] <- parse_date(df[, 1, drop = TRUE], 
                                  format = "%Y-%m-%d")
            df[, 2:9] <- map_dfc(df[, 2:9], 
                                 function(i) as.character(i))
            df[, 6:9] <- map_dfc(df[, 6:9], 
                                 function(i) {
                                   suppressWarnings(parse_double(i))
                                 })
            return(df)
          }) %>% 
  drop_na() 

colnames(dataset) <- c("Date",
                       "Manager",
                       "Category",
                       "Public",
                       "Name",
                       "InvestorsCount",
                       "MarketValue",
                       "NAV",
                       "DailyReturn")
```

```{r echo=FALSE}
dataset %>% 
  filter(Category != "Otros Fondos",
         Category != "Otros Fondos Accionarios",
         Category != "Fondo Inmobiliario de Desarrollo Nacional",
         Name != "ALIANZA RENTA FIJA MERCADOS EMERGENTES - Tipo A") %>% 
  mutate(DailyReturn = (1 + DailyReturn / 100) ^ (1 / 365) - 1) %>% 
  group_by(Category) %>% 
  summarise(MeanReturn = mean(DailyReturn) * 252,
            Volatility = sd(DailyReturn) * sqrt(252),
            Sharpe = MeanReturn / Volatility) %>% 
  arrange(desc(MeanReturn)) %>% 
  mutate(Category = str_remove(str_remove(Category, "Fondo "), "de "),
         MeanReturnT = str_c(100 * round(MeanReturn, 3), "%"),
         VolatilityT = str_c(100 * round(Volatility, 3), "%")) %>% 
  mutate(Category = parse_factor(Category)) %>% 
  filter(MeanReturn > 0) %>% 
  ggplot(mapping = aes(x = Category, 
                       y = MeanReturn,
                       fill = MeanReturn)) +
  geom_text(mapping = aes(label = MeanReturnT),
            nudge_y = 0.0075) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme(legend.position = "none",
        axis.text.x = NULL) +
  labs(y = "Mean Daily Return (APR)")
```

