---
title: "Custom visualizations in Power BI"
author: "Nicolas Cuesta Quintero"
date: 2020-02-22T21:13:14-05:00
categories: ["R"]
tags: ["Power BI", "R", "Plotly"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```


Microsoft Power BI is the current leader in the Gartner Magic Quadrant for Business Intelligence, and one of its main features is its capability for reading both R and Python code. But here is the thing, usually R plots remain as static images and lack of interactivity. Say, you want to know a point in an R scatter plot what value represents exactly. In R, you can use plotly to add this characteristic to your graphs but using plotly with Power BI default R code compiler leads to an error. The aim of this post is to show you how to embed HTML files in Power BI using R opening the doors for other libraries such as plotly, ggmap and gganimate, among others.

We will use the following libraries:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
```

## R and Power BI default visuals

### How to insert R code into Power BI 

Let's start with iris data set.

### Limitations 

## Creating custom visualizations

### Setting up your computer

### Fixing bugs

### Go for it!
