---
title: "General Binomial Models and the Cox-Ross-Rubenstein Model"
author: "Nicolas Cuesta Quintero"
date: 2020-10-01T20:00:00-05:00
categories: ["R"]
tags: ["Options pricing", "trading", "quantfinance"]
---

Pricing derivative securities has been one of the main financial problems for
both academics and practitioners. This is the price to pay for options representing
insurance claims, interest or exchange rate hedging swaps, forward prices of commodities or 
even stock options for top managerial position salaries. Usually the binomial model for
pricing derivatives at Undergraduate level is used to show how to price options by taking
the expected present value of future claims over risk-neutral probabilities. The aim of this 
post is to show a generalized version that allows to use binomial models to price any 
kind of derivative, the importance of the No Arbitrage Principle and its equivalence
to the risk-neutral probability measure (martingale), to conclude with the 
Cox-Ross-Rubinstein model for European call options. You can expect to find a summary of
the main theory behind these concepts and functions to make this practical knowledge in R.

We will use the following libraries:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
#library(DiagrammeR)
library(vtree)
```

```{r}
vtree(FakeData, "Category Group", pngknit = FALSE)
```


## Single-step asset pricing models

Let me exemplify the main point with a single-step model of a European Call option.

### European Call Options

### No Arbitrage Principle

### Risk-neutral probability measures

## The binomial model

```{r}
dvt_call <- function(s) max(s - 105, 0)

# path-independent
dvt_price <- function(h, s, u, d, r, N) {
  q <- (r - d) / (u - d)
  scenarios <- choose(N, 0:N) * q ^ (0:N) * (1 - q) ^ (N:0) * sapply(
    s * (1 + u) ^ (0:N) * (1 + d) ^ (N:0), h)
  sum(scenarios) * (1 + r) ^ (-N)
}

dvt_price(dvt_call, 100, 0.2, -0.1, 0.1, 3)
```
https://rich-iannone.github.io/DiagrammeR/#features



## The Cox-Ross-Rubenstein model

## References

[1] Capiński, M., & Kopp, E. (2012). *Discrete models of financial markets* (Vol. 1). Cambridge University Press.